---
title: "Open Syndrome workshop"
subtitle: "Filtering outpatient data with Open Syndrome Definitions, in R"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    df-print: kable
execute:
  echo: true
---

Filter real ambulatory care data with an [Open Syndrome Definition](https://opensyndrome.org),
using the [`opensyndrome`](https://github.com/OpenSyndrome/open-syndrome-r) R package.

**The four moving parts:** a dataset, a *mapping* (dataset columns → OSD concepts),
a *definition* (JSON), and the engine that turns the definition into filters.

Two chunks are yours to edit: the mapping and the definition. Run the rest in order.

```{r}
#| eval: false
# Run once
install.packages(c("dplyr", "ggplot2", "jsonlite", "knitr", "nanoparquet", "remotes", "tibble", "tidyr", "yaml"))
remotes::install_github("OpenSyndrome/open-syndrome-r")
```

```{r}
#| label: setup
#| message: false
library(opensyndrome)
library(dplyr)
library(ggplot2)
library(nanoparquet)
```

## 1. The data

Outpatient procedures from SIA/SUS, Brazil's national ambulatory care system, for the
state of Pernambuco. One row per record: an ICD-10 code, who, where and when.

SIA writes codes without the dot (`A929`) and OSD definitions use `A92.9`, so we add
`icd10_dotted` to bridge the two.

```{r}
#| label: data
sia <- read_parquet("workshop_sia_bi.parquet") |>
  mutate(
    icd10_dotted = ifelse(
      nchar(icd10) > 3,
      paste0(substr(icd10, 1, 3), ".", substr(icd10, 4, nchar(icd10))),
      icd10
    )
  )

sprintf(
  "%s records, %s columns, %s to %s",
  format(nrow(sia), big.mark = ","), ncol(sia),
  min(sia$attendance_month), max(sia$attendance_month)
)
```

```{r}
head(sia, 10)
```

::: {.callout-note collapse="true"}
## Where this data comes from

DATASUS SIASUS, group `BI` (*Boletim de Produção Ambulatorial Individualizado*) — the
patient-level outpatient record of Brazil's Unified Health System. Collected with
[PySUS](https://github.com/AlertaDengue/PySUS) from the DATASUS public FTP for the 48
processing months of 2014-01 to 2017-12, the years of the Zika epidemic, and published as
[anapaulagomes/sia-pe-2014-2017](https://huggingface.co/datasets/anapaulagomes/sia-pe-2014-2017).

**Curation.** De-identified upstream: no patient card number, birth date reduced to a year,
ethnicity removed. Sex codes decoded to `Female` / `Male`, the six-digit DATASUS municipality
code resolved to a city and a state through the IBGE API, `YYYYMM` dates to `YYYY-MM`.
DATASUS stores age as a number plus a unit — days and months both mean under a year, and a
separate unit counts the years past 100 — all collapsed into whole years, then bucketed into
`age_group`: Infants under 1, Children and adolescents 1-15, Adults 16-59, Elderly 60+.
DATASUS files carry no real nulls, only fill codes, so the `0000` diagnosis fill, the `99`
ignored-age sentinel and unusable municipality codes were read as missing and their records
dropped: 6.6M of the 17.6M raw records survive, the absent diagnosis alone accounting for 10.7M.

**Read it carefully.** A row is a procedure billed to the SUS, not a confirmed case, and one
patient can appear many times. The finest time resolution is the month, so there is no
epidemiological week. Billing is retroactive, which is why 8,435 rows have an attendance month
back in 2013. `city` is the patient's municipality of residence, not where they were seen:
27 states appear, though 98.5% of the records are in Pernambuco. 6,704 distinct ICD-10 codes
are present.

**License.** The records are public data produced by the SUS under Lei 12.527/2011 and Decreto
8.777/2016; the curation layer is CC BY 4.0. Cite DATASUS for the data.
:::

## 2. Map your columns to OSD concepts

A definition talks about *concepts* — `diagnosis`, `symptom`, `demographic_criteria`,
`diagnostic_test`, `epidemiological_history` — never about your column names. The mapping
is the bridge, and `value_encodings` translates OSD's canonical values (`male`) into
whatever your data writes (`Male`).

It is a YAML file, the same format the Python package uses; we keep it inline so
everything lives in one document.

```{r}
#| label: mapping
mapping_yaml <- '
profiles:
  - name: sia_bi
    value_encodings:
      sex:
        male: "Male"
        female: "Female"
    columns:
      icd10_dotted:
        concept: diagnosis
        system: ICD-10
        dtype: string
      age_years:
        concept: demographic_criteria
        attribute: age
        dtype: integer
      sex:
        concept: demographic_criteria
        attribute: sex
        dtype: string
'

mapping_file <- tempfile(fileext = ".yaml")
writeLines(mapping_yaml, mapping_file)

profile <- os_read_profile(mapping_file, "sia_bi")
profile
```

`os_validate_profile()` catches mapping mistakes before you classify anything:
a column that does not exist, or a `dtype` the data cannot honour.

```{r}
os_validate_profile(profile, sia)
```

## 3. The definition

An OSD definition is a JSON document. Everything above `inclusion_criteria` is metadata —
who wrote it, where, for which threat — and none of it is executed. Only the criteria run:
a row has to satisfy every inclusion criterion and no exclusion criterion, and groups
inside combine with `AND`, `OR` or `AT_LEAST`.

Paste your own, or tweak this one.

```{r}
#| label: definition
definition_json <- '{
  "title": "Arboviral illness",
  "scope": "broad",
  "category": "suspected",
  "version": "1.0.0",
  "open_syndrome_version": "1.0.0",
  "published_in": "https://opensyndrome.org",
  "published_at": "2026-01-01T12:00:00Z",
  "location": "Pernambuco, Brazil",
  "language": "English",
  "organization": "Open Syndrome workshop",
  "definition_type": "syndrome_definition",
  "status": "draft",
  "human_readable_definition": "Outpatient record coded as dengue, chikungunya, Zika or another mosquito-borne viral fever.",
  "references": [{"url": "https://icd.who.int/browse10/2019/en#/A90"}],
  "inclusion_criteria": [
    {
      "type": "criterion",
      "logical_operator": "OR",
      "values": [
        {"type": "diagnosis", "name": "Dengue fever",
         "code": {"system": "ICD-10", "code": "A90"}},
        {"type": "diagnosis", "name": "Dengue haemorrhagic fever",
         "code": {"system": "ICD-10", "code": "A91"}},
        {"type": "diagnosis", "name": "Chikungunya virus disease",
         "code": {"system": "ICD-10", "code": "A92.0"}},
        {"type": "diagnosis", "name": "Zika virus disease",
         "code": {"system": "ICD-10", "code": "A92.5"}},
        {"type": "diagnosis", "name": "Other mosquito-borne viral fevers",
         "code": {"system": "ICD-10", "code": "A92.8"}},
        {"type": "diagnosis", "name": "Mosquito-borne viral fever, unspecified",
         "code": {"system": "ICD-10", "code": "A92.9"}}
      ]
    }
  ]
}'

definition <- jsonlite::fromJSON(definition_json, simplifyVector = FALSE)
os_validate(definition)
```

`os_validate()` checks the structure. For the full JSON Schema check, use
`os_validate(definition, strict = TRUE)` — that needs the `jsonvalidate` package
and `os_download_schema()`.

Two questions to ask any definition before you run it: what does it need from my
data, and what is it actually saying?

```{r}
os_required_fields(definition)
```

```{r}
os_describe(definition)
```

::: {.callout-tip collapse="true"}
## Try adding a criterion

Narrow it down to adults, using the `demographic_criteria` mapping — add this to
`values` and change the outer `logical_operator` to `"AND"`:

```json
{"type": "demographic_criteria", "name": "Adults",
 "attribute": "age", "operator": ">=", "value": 18}
```
:::

## 4. Filter

`os_label()` adds one logical column per definition, so nothing is thrown away —
you keep every row and know which ones match. (`os_filter()` returns only the
matching rows, when that is what you want.)

```{r}
#| label: label
labelled <- os_label(sia, list(arboviral_illness = definition), profile)

sprintf(
  "%s matching records (%.4f%% of the dataset)",
  format(sum(labelled$arboviral_illness), big.mark = ","),
  100 * mean(labelled$arboviral_illness)
)
```

```{r}
labelled |>
  filter(arboviral_illness) |>
  select(icd10, icd10_dotted, age_years, sex, attendance_month, city) |>
  head(10)
```

## 5. Time series of the matches

```{r}
#| label: monthly-helper
monthly_counts <- function(labelled_df, names, month_column = "attendance_month") {
  labelled_df |>
    mutate(month = as.Date(paste0(.data[[month_column]], "-01"))) |>
    summarise(across(all_of(names), sum), .by = month) |>
    arrange(month) |>
    tidyr::pivot_longer(all_of(names), names_to = "definition", values_to = "matches")
}

plot_monthly <- function(counts) {
  ggplot(counts, aes(month, matches, colour = definition)) +
    geom_line() +
    geom_point(size = 1) +
    scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
    expand_limits(y = 0) +
    labs(x = "Month of attendance", y = "Matching records", colour = NULL) +
    theme_minimal() +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r}
#| label: timeseries
#| fig-height: 3.5
monthly_counts(labelled, "arboviral_illness") |> plot_monthly()
```

## 6. Definitions from the community repository

Definitions are shared in [OpenSyndrome/definitions](https://github.com/OpenSyndrome/definitions).
`os_download_definitions()` mirrors them into `~/.open_syndrome/v1` (the same cache the
Python package uses) and `os_list()` searches what you have. Same data, same mapping, same
engine — only the definitions change, and they were written by other people for other places.

```{r}
#| label: download
#| message: false
# Download once; skip it when the cache is already populated
if (nrow(os_list()) == 0) os_download_definitions()

catalogue <- lapply(c("arbovirosis", "arbovirus", "zika", "dengue"), \(term) os_list(disease = term)) |>
  bind_rows() |>
  distinct()

catalogue |> select(name, title, disease, location, organization)
```

`os_load()` takes a name from the cache, a local file, or a URL.

```{r}
#| label: repo-label
repo_definitions <- setNames(lapply(catalogue$name, os_load), catalogue$name)

repo_labelled <- os_label(sia, repo_definitions, profile)

repo_summary <- tibble::tibble(
  definition = catalogue$name,
  location = catalogue$location,
  matches = sapply(catalogue$name, \(name) sum(repo_labelled[[name]]))
) |>
  arrange(desc(matches))

repo_summary
```

```{r}
#| label: repo-timeseries
#| fig-height: 3.5
monthly_counts(repo_labelled, repo_summary$definition) |> plot_monthly()
```

::: callout-important
## Why three definitions match nothing

Look at the warnings above: *"No column mapped to concept 'symptom'"*.

`zika_paho`, `denguefever_malaysia` and `arbovirosis_paraguay_cd` are written in terms
of symptoms (rash, arthralgia, fever) or clinical judgement. This dataset has ICD-10
codes and nothing else, so the engine has no column to evaluate them against and skips
those criteria — by default with a warning; pass `on_unresolvable = "error"` to stop
instead.

Same syndrome, different data source, different definition. That is the problem OSD
exists to make explicit.

```{r}
os_describe(repo_definitions$zika_paho)
```
:::
